---
title: "Census data: aggregation queries"
format: html
---

Testing DuckDB queries against our census tables — the same queries
we'll run in the frontend via DuckDB-WASM.

**Part 1**: Arrondissement-level demographics (`census_population.parquet`)
**Part 2**: DA-level star schema, multi-year (`geo_da`, `census_demographics`, `census_economics`, `census_income`)

```{r}
library(duckdb)
con <- dbConnect(duckdb())
```

# Part 1: Arrondissement-level demographics

## Load the parquet file

```{r}
parquet_path <- here::here("pipelines", "transform", "input", "census_population.parquet")
dbExecute(con, sprintf("CREATE VIEW pop AS SELECT * FROM '%s'", parquet_path))
```

### Quick look at the data

```{r}
dbGetQuery(con, "SHOW ALL TABLES")
```

```{r}
dbGetQuery(con, "SELECT * FROM pop")
```

```{r}
dbGetQuery(con, "
  SELECT year, source, sex,
         COUNT(DISTINCT arrondissement) AS districts,
         COUNT(*) AS rows
  FROM pop
  GROUP BY year, source, sex
  ORDER BY year, source, sex
")
```

## 1. Total population by district (2021)

The simplest aggregation — sum across all age groups for sex='total'.

```{r}
dbGetQuery(con, "
  SELECT arrondissement, SUM(population) AS total_pop
  FROM pop
  WHERE year = 2021 AND sex = 'total'
  GROUP BY arrondissement
  ORDER BY total_pop DESC
")
```

## 2. Age pyramid for a single district

Classic demographic query — male/female counts by age group for one arrondissement.

```{r}
dbGetQuery(con, "
  SELECT age_group, sex, population
  FROM pop
  WHERE arrondissement = 'Le Plateau-Mont-Royal'
    AND year = 2021
    AND sex IN ('male', 'female')
  ORDER BY age_group, sex
")
```

## 3. Compare age distribution across districts

Which districts are youngest/oldest? Compute share of population under 15 and over 65.

```{r}
dbGetQuery(con, "
  WITH totals AS (
    SELECT arrondissement, SUM(population) AS total
    FROM pop WHERE year = 2021 AND sex = 'total'
    GROUP BY arrondissement
  ),
  young AS (
    SELECT arrondissement, SUM(population) AS pop_young
    FROM pop
    WHERE year = 2021 AND sex = 'total'
      AND age_group IN ('0-4', '5-9', '10-14')
    GROUP BY arrondissement
  ),
  old AS (
    SELECT arrondissement, SUM(population) AS pop_old
    FROM pop
    WHERE year = 2021 AND sex = 'total'
      AND age_group IN ('65-69', '70-74', '75-79', '80-84', '85-89', '90-94', '95-99', '100+')
    GROUP BY arrondissement
  )
  SELECT t.arrondissement,
         t.total,
         ROUND(100.0 * y.pop_young / t.total, 1) AS pct_under_15,
         ROUND(100.0 * o.pop_old / t.total, 1) AS pct_over_65,
         ROUND(100.0 * (t.total - y.pop_young - o.pop_old) / t.total, 1) AS pct_working_age
  FROM totals t
  JOIN young y USING (arrondissement)
  JOIN old o USING (arrondissement)
  ORDER BY pct_over_65 DESC
")
```

## 4. Population change 2016 → 2021

Compare total population across census years for districts that have data in both.

```{r}
dbGetQuery(con, "
  WITH by_year AS (
    SELECT arrondissement, year, SUM(population) AS total
    FROM pop
    WHERE sex = 'total'
    GROUP BY arrondissement, year
  )
  SELECT a.arrondissement,
         a.total AS pop_2016,
         b.total AS pop_2021,
         b.total - a.total AS change,
         ROUND(100.0 * (b.total - a.total) / a.total, 1) AS pct_change
  FROM by_year a
  JOIN by_year b ON a.arrondissement = b.arrondissement
  WHERE a.year = 2016 AND b.year = 2021
  ORDER BY pct_change DESC
")
```

## 5. Sex ratio by district (2021)

Male/female ratio — only for districts with male/female breakdown.

```{r}
dbGetQuery(con, "
  WITH by_sex AS (
    SELECT arrondissement, sex, SUM(population) AS total
    FROM pop
    WHERE year = 2021 AND sex IN ('male', 'female')
    GROUP BY arrondissement, sex
  )
  SELECT m.arrondissement,
         m.total AS male,
         f.total AS female,
         ROUND(100.0 * m.total / (m.total + f.total), 1) AS pct_male
  FROM by_sex m
  JOIN by_sex f ON m.arrondissement = f.arrondissement
  WHERE m.sex = 'male' AND f.sex = 'female'
  ORDER BY pct_male DESC
")
```

## 6. Data coverage check

Which districts have male/female breakdown vs total-only?
This matters for 2016 arrondissements (Ville de Montréal source = total only).

```{r}
dbGetQuery(con, "
  SELECT year, source,
         COUNT(DISTINCT CASE WHEN sex = 'male' THEN arrondissement END) AS has_sex_split,
         COUNT(DISTINCT CASE WHEN sex = 'total' THEN arrondissement END) AS has_total,
         COUNT(DISTINCT arrondissement) AS all_districts
  FROM pop
  GROUP BY year, source
  ORDER BY year, source
")
```

## 7. Dependency ratio

Classic demographic indicator: (population < 15 + population 65+) / population 15-64.
Only meaningful with total population (not sex-split to avoid double-counting).

```{r}
dbGetQuery(con, "
  WITH classified AS (
    SELECT arrondissement, year,
           CASE
             WHEN age_group IN ('0-4','5-9','10-14') THEN 'young'
             WHEN age_group IN ('65-69','70-74','75-79','80-84','85-89','90-94','95-99','100+') THEN 'old'
             ELSE 'working'
           END AS age_class,
           population
    FROM pop
    WHERE sex = 'total'
  )
  SELECT arrondissement, year,
         SUM(CASE WHEN age_class = 'young' THEN population ELSE 0 END) AS young,
         SUM(CASE WHEN age_class = 'working' THEN population ELSE 0 END) AS working,
         SUM(CASE WHEN age_class = 'old' THEN population ELSE 0 END) AS old,
         ROUND(100.0 * (
           SUM(CASE WHEN age_class IN ('young','old') THEN population ELSE 0 END)
         ) / NULLIF(SUM(CASE WHEN age_class = 'working' THEN population ELSE 0 END), 0), 1) AS dependency_ratio
  FROM classified
  GROUP BY arrondissement, year
  HAVING SUM(population) > 1000
  ORDER BY year, dependency_ratio DESC
")
```

# Part 2: DA-level star schema

The DA-level data uses a star schema: one geometry table (`geo_da`) joined with
long-format fact tables. This is the architecture for the DuckDB-WASM frontend.

## Load the star schema tables

```{r}
dbExecute(con, "INSTALL spatial; LOAD spatial;")

load_dir <- here::here("pipelines", "load", "output")
dbExecute(con, sprintf("CREATE VIEW geo AS SELECT * FROM '%s/geo_da.parquet'", load_dir))
dbExecute(con, sprintf("CREATE VIEW demo AS SELECT * FROM '%s/census_demographics.parquet'", load_dir))
dbExecute(con, sprintf("CREATE VIEW econ AS SELECT * FROM '%s/census_economics.parquet'", load_dir))
dbExecute(con, sprintf("CREATE VIEW income AS SELECT * FROM '%s/census_income.parquet'", load_dir))
```

### Table sizes (by year)

```{r}
dbGetQuery(con, "
  SELECT 'geo_da' AS tbl, year, COUNT(*) AS rows FROM geo GROUP BY year
  UNION ALL SELECT 'demographics', year, COUNT(*) FROM demo GROUP BY year
  UNION ALL SELECT 'economics', year, COUNT(*) FROM econ GROUP BY year
  UNION ALL SELECT 'income', year, COUNT(*) FROM income GROUP BY year
  ORDER BY tbl, year
")
```

## 8. Age pyramid at DA level (2021)

Same query pattern as arrondissement, but now at dissemination area granularity.

```{r}
dbGetQuery(con, "
  SELECT age_group, sex, population
  FROM demo
  WHERE geo_uid = '24660101' AND year = 2021 AND sex != 'total'
  ORDER BY age_group, sex
")
```

## 9. Youngest and oldest DAs (2021)

Which DAs have the highest share of seniors (65+)?

```{r}
dbGetQuery(con, "
  WITH totals AS (
    SELECT geo_uid, SUM(population) AS total
    FROM demo WHERE sex = 'total' AND year = 2021
    GROUP BY geo_uid
    HAVING total > 100
  ),
  old AS (
    SELECT geo_uid, SUM(population) AS pop_65plus
    FROM demo
    WHERE sex = 'total' AND year = 2021
      AND age_group IN ('65-69','70-74','75-79','80-84','85-89','90-94','95-99','100+')
    GROUP BY geo_uid
  )
  SELECT t.geo_uid,
         t.total,
         o.pop_65plus,
         ROUND(100.0 * o.pop_65plus / t.total, 1) AS pct_65plus
  FROM totals t JOIN old o USING (geo_uid)
  ORDER BY pct_65plus DESC
  LIMIT 15
")
```

## 10. Economics: variable overview (by year)

What variables are available in the economics table?

```{r}
dbGetQuery(con, "
  SELECT year, variable, sex, COUNT(*) AS n_das, ROUND(AVG(value), 1) AS avg_val
  FROM econ
  GROUP BY year, variable, sex
  ORDER BY year, variable, sex
")
```

## 11. Median household income — top and bottom DAs

```{r}
dbGetQuery(con, "
  SELECT geo_uid, value AS median_hh_income
  FROM econ
  WHERE variable = 'median_income_household'
    AND value IS NOT NULL
  ORDER BY value DESC
  LIMIT 10
")
```

```{r}
dbGetQuery(con, "
  SELECT geo_uid, value AS median_hh_income
  FROM econ
  WHERE variable = 'median_income_household'
    AND value IS NOT NULL AND value > 0
  ORDER BY value ASC
  LIMIT 10
")
```

## 12. Low income prevalence (LIM-AT)

Which DAs have the highest share of population in low income?

```{r}
dbGetQuery(con, "
  SELECT geo_uid, value AS lim_at_pct
  FROM econ
  WHERE variable = 'lim_at_prevalence'
    AND value IS NOT NULL
  ORDER BY value DESC
  LIMIT 15
")
```

## 13. Income distribution: histogram for one DA

The full bracket breakdown — what you'd display as a bar chart on click.

```{r}
dbGetQuery(con, "
  SELECT bracket, count
  FROM income
  WHERE geo_uid = '24660101'
    AND income_type = 'total_income'
    AND sex = 'total'
  ORDER BY count DESC
")
```

## 14. Income inequality across DAs

Compute % earning over $100K from the bracket data.

```{r}
dbGetQuery(con, "
  WITH brackets AS (
    SELECT geo_uid,
      SUM(count) AS total_pop,
      SUM(CASE WHEN bracket IN ('$100,000-$149,999', '$150,000+')
               THEN count ELSE 0 END) AS over_100k
    FROM income
    WHERE income_type = 'total_income' AND sex = 'total'
    GROUP BY geo_uid
    HAVING total_pop > 100
  )
  SELECT geo_uid, total_pop, over_100k,
         ROUND(100.0 * over_100k / total_pop, 1) AS pct_over_100k
  FROM brackets
  ORDER BY pct_over_100k DESC
  LIMIT 15
")
```

## 15. Join pattern: geo + economics (with year)

The fundamental frontend query — join geometry with a fact table for choropleths.
Note: must join on both `geo_uid` AND `year` since DA boundaries differ between censuses.

```{r}
dbGetQuery(con, "
  SELECT g.geo_uid, g.area_sqkm, e.value AS median_income
  FROM geo g
  JOIN econ e ON g.geo_uid = e.geo_uid AND g.year = e.year
  WHERE e.variable = 'median_income_household' AND e.year = 2021
  ORDER BY e.value DESC
  LIMIT 10
")
```

## 16. Gender income gap by DA (2021)

Compare male vs female median income.

```{r}
dbGetQuery(con, "
  WITH by_sex AS (
    SELECT geo_uid, sex, value
    FROM econ
    WHERE variable = 'median_income' AND year = 2021
      AND sex IN ('male', 'female')
      AND value IS NOT NULL AND value > 0
  )
  SELECT m.geo_uid,
         m.value AS male_median,
         f.value AS female_median,
         ROUND(100.0 * (m.value - f.value) / m.value, 1) AS gap_pct
  FROM by_sex m
  JOIN by_sex f ON m.geo_uid = f.geo_uid
  WHERE m.sex = 'male' AND f.sex = 'female'
  ORDER BY gap_pct DESC
  LIMIT 15
")
```

## 17. Household income distribution comparison: renters vs owners (2021)

Cross-table query: join economics (tenure) with income brackets for DAs
that are majority-renter vs majority-owner.

```{r}
dbGetQuery(con, "
  WITH tenure AS (
    SELECT geo_uid,
      MAX(CASE WHEN variable = 'tenure_owner' THEN value END) AS owners,
      MAX(CASE WHEN variable = 'tenure_renter' THEN value END) AS renters
    FROM econ
    WHERE variable IN ('tenure_owner', 'tenure_renter') AND year = 2021
    GROUP BY geo_uid
    HAVING owners > 0 AND renters > 0
  ),
  classified AS (
    SELECT geo_uid,
      CASE WHEN renters > owners THEN 'majority_renter' ELSE 'majority_owner' END AS tenure_type
    FROM tenure
  )
  SELECT c.tenure_type, i.bracket,
         SUM(i.count) AS total_count
  FROM classified c
  JOIN income i ON c.geo_uid = i.geo_uid
  WHERE i.income_type = 'household_total_income' AND i.sex = 'total'
  GROUP BY c.tenure_type, i.bracket
  ORDER BY c.tenure_type, i.bracket
")
```

# Part 3: Cross-year comparison (2011 vs 2021)

## 18. Population change by DA (2011 vs 2021)

Compare total population across census years. Only DAs that exist in both years
(same geo_uid — note boundaries may have changed slightly).

```{r}
dbGetQuery(con, "
  WITH by_year AS (
    SELECT year, geo_uid, value AS pop
    FROM econ
    WHERE variable = 'pop_total'
  )
  SELECT a.geo_uid,
         a.pop AS pop_2011,
         b.pop AS pop_2021,
         b.pop - a.pop AS change,
         ROUND(100.0 * (b.pop - a.pop) / NULLIF(a.pop, 0), 1) AS pct_change
  FROM by_year a
  JOIN by_year b ON a.geo_uid = b.geo_uid
  WHERE a.year = 2011 AND b.year = 2021
  ORDER BY pct_change DESC
  LIMIT 15
")
```

## 19. Language shift: English vs French mother tongue

How did the English/French mother tongue balance change at DA level?

```{r}
dbGetQuery(con, "
  WITH lang AS (
    SELECT year, geo_uid, variable, value
    FROM econ
    WHERE variable IN ('lang_mother_english', 'lang_mother_french')
      AND sex = 'total'
  )
  SELECT year,
         SUM(CASE WHEN variable = 'lang_mother_english' THEN value END) AS english,
         SUM(CASE WHEN variable = 'lang_mother_french' THEN value END) AS french,
         ROUND(100.0 * SUM(CASE WHEN variable = 'lang_mother_english' THEN value END)
           / NULLIF(SUM(CASE WHEN variable = 'lang_mother_english' THEN value END)
             + SUM(CASE WHEN variable = 'lang_mother_french' THEN value END), 0), 1) AS pct_english
  FROM lang
  GROUP BY year
  ORDER BY year
")
```

## 20. Age structure comparison (2011 vs 2021)

Compare the share of young (0-14) and seniors (65+) across years.
Note: 2011 top bucket is '85+', 2021 has 85-89, 90-94, 95-99, 100+.

```{r}
dbGetQuery(con, "
  WITH classified AS (
    SELECT year,
      CASE
        WHEN age_group IN ('0-4','5-9','10-14') THEN 'young'
        WHEN age_group IN ('65-69','70-74','75-79','80-84') THEN 'old'
        WHEN age_group IN ('85+','85-89','90-94','95-99','100+') THEN 'old'
        ELSE 'working'
      END AS age_class,
      population
    FROM demo
    WHERE sex = 'total'
  )
  SELECT year,
    SUM(population) AS total,
    ROUND(100.0 * SUM(CASE WHEN age_class = 'young' THEN population END)
      / SUM(population), 1) AS pct_young,
    ROUND(100.0 * SUM(CASE WHEN age_class = 'old' THEN population END)
      / SUM(population), 1) AS pct_65plus,
    ROUND(100.0 * SUM(CASE WHEN age_class = 'working' THEN population END)
      / SUM(population), 1) AS pct_working
  FROM classified
  GROUP BY year
  ORDER BY year
")
```

```{r}
dbDisconnect(con)
```
